Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2015-10-22 RMS/submission.tex
temporary commit

\section{Introduction}
Periodic task scheduling is one of the most important topics within
the field of real-time systems, due to the large number of control
systems that require cyclic activities~\cite{buttazzo2011hard}. A set
of periodic tasks is said to be \emph{schedulable} with respect to
some scheduling algorithm if all jobs meet their
deadlines. \emph{Rate-Monotonic Scheduling} (\emph{RMS}) is a
\emph{fixed} priority scheduling algorithm for preemptive hard
real-time environments proposed by Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73}, which assigns priorities to
jobs according to the periods of the corresponding tasks: the smaller
period, the higher priority. It is proved to be the \emph{optimal}
fixed priority scheduling algorithm~\cite{DBLP:journals/jacm/LiuL73},
in the sense that any set of tasks, which is schedulable under
\emph{some} fixed priority scheduling algorithm, is also schedulable
with respect to RMS. It is widely used in safety-critical real-time
applications, such as trains and avionics, thanks to its optimality
and easiness to implement.

@@

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
modeling language and analyzing tool for real-time systems, to model a
realistic implementation of RMS algorithm and then verify some desired
properties on the built model. Based on a realistic implementation,
our model extends the original ideal one proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by allowing computation
requirements of tasks to be non-constant, and by considering overhead
and other details of the enviroments.

The rest of this paper is organized as follows. Section~\ref{s:background}
introduces the standard setting of the RMS algorithm proposed
in~\cite{DBLP:journals/jacm/LiuL73} and our
implementation. Section~\ref{s:imp} gives a background to
Real-Time Maude. Section~\ref{s:formalism} presents how we model the
RMS implementation using Real-Time Maude. Then
Section~\ref{s:verification} explains how to verify the desired
properties and to analyze the results. Related work is discussed in
Section~\ref{s:relate}. We conclude the paper in
Section~\ref{s:conclusion}.

@@

\subsection{Rate-Monotonic Scheduling Algorithm}
\label{ss:rms}
%\subsection{Formulation of the Standard Setting}
A task set consists of \emph{only} $n$ periodic tasks
$\tau_1,\ldots,\tau_n$. Task $\tau_i$ has a period $T_i$ and a
computation requirement $C_i$. First jobs of all tasks are assumed to
be initiated simultaneously. This means that jobs corresponding to
task $\tau_i$ are initiated at times $kT_i$ with integers $k\ge
0$. Deadlines consist of runability constraints only: the job
initiated at time $kT_i$ has $(k+1)T_i$ as its deadline, which is the
initiation time of the next job. The RMS algorithm chooses the
labeling such that $T_1\le T_2\le \ldots \le T_n$. Consequently,
$\tau_i$ receives priority $i$, assuming smaller numbers have higher
priorities. The following assumptions are made:

@@

However in this paper, we consider an implementation instead of the
RMS algorithm itself, thus the model would be more complicated than
this ideal setting.  Assumptions~(A1) and (A3) will be relaxed later
to obtain a more realisitc analysis model.

@@

\subsubsection{Specification}
Real-Time Maude models a system using \emph{modules}. A module
specifies a \emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup
A , IR, TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}. The function symbols are allowed to be mixfix.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of conditional equations on $\Sigma$, and
  $A$ a set of equational axioms such as associativiy, commutativity
  and identity.  $(\Sigma, E\cup A)$ models the system's ``static''
  states as an algebraic data type, and includes a built-in
  specification of a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~t\rightarrow t'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instaneous transition},
  without consuming time, from an instance of $t$ to the corresponding
  instance of $t'$, \emph{provided} the conditions hold.
\item $TR$ is a set of (\emph{labeled}) \emph{tick rules}
  $[l]~:~\{t\}\rightarrow\{t'\} \mbox{ \textbf{in time} }r\mbox{
  \textbf{if} }cond$ that specify \emph{timed transitions} advancing
  time in the \emph{entire} state $t$ by $r$ time units. $IR$ and $TR$
  together model the ``dynamic'' behaviors of the system.
\end{itemize}

@@

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic operating system.
Interrupts triggered by the clock would come every $T$, which we call
\emph{interrupt cycle}. When an interrupt request occurs, if the
system is interruptable, i.e. the interrupt mask is cleared, the
handler function $schedule()$ will be invoked; otherwise $schedule()$
will be pending until the interrupt mask becomes cleared.  The
implementation is shown as $schedule()$ in Algorithm~\ref{a:schedule},
where $taskList$ is the list of periodic tasks to be scheduled. We assume that 
the list is ordered descendingly by priorities, and both variables
$taskList$ and $timer$ are global. In this implementation, there is
only one kind of interrupt, the period $T_i$ of each task is a
multiple of $T$, and the tasks are independent, meeting
assumption~(A2).

@@

In Algorithm~\ref{a:schedule}, the handler function $scheduler()$
first updates status of all tasks in $taskList$ via function
$updateStatus()$. Then it scans the list to execute the ready tasks
one by one, or to do a return when encountering an interrupted task.
The function $updateStatus()$ updates each task in two stages:
firstly, if the task is running, it becomes interrupted\footnote{Note
  that the status \textit{INTERRUPT} indicates the task is interrupted
  for the moment, or was interrupted before but its execution is not
  finished yet.}; secondly for the task at its initiation time, if its
previous job is finished, it would be set to be ready, otherwise it
produces an error. Notice that $schedule()$ is invoked only when the
periodic interrupt is handled. Its running cannot be interrupted when
it is updating status of tasks or finding the next task to be
executed, however, it can be interrupted while executing some task
(Line~\ref{l:function}). This allows the running of $schedule()$ to be
nested.

In the rest of this paper, we use \emph{scheduling} to refer to the
stage, from the moment when a pending interrupt request is detected,
to the moment when the first should-be-run periodic task starts
executing, i.e. Line~\ref{l:function} in
Algorithm~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when handling interrupt
requests, (ii) the time spent by $schedule()$ finding and setting the
\emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}), and (iii) the time for switching context
from $schedule()$ to that first should-be-run task. \emph{Switching}
refers to the stage, from the moment when a periodic task completes
its execution, to the moment when the next should-be-run periodic task
starts executing. \emph{Switching time} thus also consists of three
parts: (i) the time for switching context from the finished task back
to $schedule()$, (ii) the time spent by $schedule()$ finding and
setting the \emph{next} should-be-run periodic task, and (iii) the
time for switching context from $schedule()$ to that next
should-be-run task.

@@

These assumptions make our model different from the standard one. For
instance, if an interrupt request occurs while switching is performed,
jobs corresponding to task $\tau_i$ could not be initiated at exactly
$kT_i$. They should wait until switching is finished and the interrupt
mask is cleared. Note that (A2) says that jobs are ready to run at
their initiation times, however, no one can start running at exactly
its initiation time, because scheduling takes time.

\jx{TODO. A figure is needed here to demostrate the scheduling under
  the three assumptions above.}

In this section, we introduce first how we model the states--the
static aspect--of the system using algebraic data types, then how we
specify the essential behaviors--the dynamic aspect--using rewrite
rules.

\subsection{Basic Data Types}
In our model, tasks are identified with their indexes of sort
\verb|Nat| in the $taskList$. We define a sort \verb|MaybeNat|
wrapping \verb|Nat|s to refer to some task, with \emph{constructor}
\verb|some| followed by a \verb|Nat| $n$ indicating the task with
index $n$, and \verb|none| for no task:
\begin{verbatim}
  op none : -> MaybeNat [ctor] .
  op some_ : Nat -> MaybeNat [ctor] .
\end{verbatim}

@@

We also need a sort \verb|Interval| denoting the interval, within
which the computation time of a task varies and which we call
\emph{possible finishing interval}, and a sort \verb|Counter| to
represent the time some task has been running and the corresponding
possible finishing interval.
\begin{verbatim}
  op [_,_] : Time Time -> Interval [ctor] .
  op [_/_] : Time Interval -> Counter [ctor] .
\end{verbatim}

@@

\subsection{Interrupt Requests}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instaneous action, and thus is modeled by the
following instaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Algorithm~\ref{a:schedule}, which is modeled by
the recursive operation as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and \verb|update_with_| updating the status of individual task
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Algorithm~\ref{a:schedule})
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, status : ST > with TIMER
        = if ST == DORMANT then < O : PTask | status : READY >
          else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > with TIMER
       = if ST == RUNNING then < O : PTask | status : INTERRUPT >
         else < O : PTask |> fi [owise] .
\end{verbatim}
In the case where \verb|TIMER|($timer$) can be divided by the task's
period, i.e. the task should be initiated, if the status is
\verb|DORMANT|, it should be ready to run; otherwise it means the
previous job of the task is not finished, thus it misses its deadline,
producing an \verb|error|. In the other case where the tasks should
not be initiated, the status of the task changes only if it is
\verb|RUNNING|. We can see that \verb|updateStatus_with_| behaves the
same as $updateStatus()$.

@@

\subsection{Timed Behaviors of the System}
\emph{Timed behaviors} of the system consist of two parts: the
execution of tasks and the execution of the interrupt source. Both are
modeled into the following one ``standard'' tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instaneous
transition \emph{must} be performed. Notice that the variable $R$ is
\emph{continuous} with respect to the specific time
domain\footnote{Real-Time Maude contains built-in modules to define
  the time domain to be natural numbers and rational numbers,
  specifying \emph{discrete} time domains and \emph{dense} time
  domains respectively.}  that we choose to instantiate the model on,
which is different from timed automata that discretize dense time by
defining ``clock region''.

>>>>

\section{Introduction}
Periodic task scheduling is one of the most important topics within
the field of real-time systems, due to the large number of control
systems that require cyclic activities~\cite{buttazzo2011hard}. A set
of periodic tasks is said to be \emph{schedulable} with respect to
some scheduling algorithm if all jobs meet their
deadlines. \emph{Rate-Monotonic Scheduling} (\emph{RMS}) is a
\emph{fixed} priority scheduling algorithm for preemptive hard
real-time environments proposed by Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73}, which assigns priorities to
jobs according to the periods of the corresponding tasks: the smaller
period, the higher priority. RMS is proved to be the \emph{optimal}
fixed priority scheduling algorithm~\cite{DBLP:journals/jacm/LiuL73},
in the sense that any set of tasks, which is schedulable under
\emph{some} fixed priority scheduling algorithm, is also schedulable
with respect to RMS. It is widely used in safety-critical real-time
applications, such as trains and avionics, thanks to its optimality
and easiness to implement.

@@

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
modeling language and analyzing tool for real-time systems, to model a
realistic implementation of RMS and then verify some desired
properties on the built model. Based on a realistic implementation,
our model extends the standard one proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by allowing computation
requirements of tasks to be non-constant, and by considering overhead
and other details of the enviroments.

The rest of this paper is organized as
follows. Section~\ref{s:background} gives some background of the RMS
algorithm and Real-Time Maude.  Section~\ref{s:imp} presents the RMS
implementation that we model and analyze.  Section~\ref{s:formalism}
introduces how we model the RMS implementation using Real-Time
Maude. Then Section~\ref{s:verification} explains how to verify the
desired properties and to evaluate the results. Related work is
discussed in Section~\ref{s:relate}. We conclude the paper in
Section~\ref{s:conclusion}.

@@

\subsection{Rate-Monotonic Scheduling Algorithm}
\label{ss:rms}
%\subsection{Formulation of the Standard Setting}
A task set consists of \emph{only} $n$ periodic tasks
$\tau_1,\ldots,\tau_n$. Each task $\tau_i$ has a period $T_i$ and a
computation requirement $C_i$. First jobs of all tasks are assumed to
be initiated simultaneously. This means that jobs corresponding to
task $\tau_i$ are initiated at times $kT_i$ with integers $k\ge
0$. Deadlines consist of runability constraints only: the job
initiated at time $kT_i$ has $(k+1)T_i$ as its deadline, which is the
initiation time of the next job. The RMS algorithm chooses the
labeling such that $T_1\le T_2\le \ldots \le T_n$. Consequently,
$\tau_i$ receives priority $i$, assuming smaller numbers have higher
priorities. The following assumptions are made:

@@

However in this paper, we consider an implementation instead of the
RMS algorithm itself, thus the model would be more complicated than
this standard, ideal setting.  Assumptions~(A1) and (A3) will be
relaxed later to obtain a more realistic analysis model.

@@

\subsubsection{Specification}
Real-Time Maude models systems as \emph{modules}. A module specifies a
\emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup A , IR,
TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}. The function symbols are allowed to be mixfix.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of conditional equations on $\Sigma$, and
  $A$ a set of equational axioms such as associativity, commutativity
  and identity.  $(\Sigma, E\cup A)$ models the system's ``static''
  states as an algebraic data type, and includes a built-in
  specification of a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~t\rightarrow t'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instaneous transition},
  without consuming time, from an instance of $t$ to the corresponding
  instance of $t'$, \emph{provided} the conditions hold.
\item $TR$ is a set of (\emph{labeled}) \emph{tick rules}
  $[l]~:~\{t\}\rightarrow\{t'\} \mbox{ \textbf{in time} }r\mbox{
  \textbf{if} }cond$ that specify \emph{timed transitions} advancing
  time in the \emph{entire} state $t$ by $r$ time units. $IR$ and $TR$
  together model the ``dynamic'' behaviors of the system.
\end{itemize}

@@

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic operating system.
Interrupts would be triggered by the clock every $T$, which we call
\emph{interrupt cycle}. When an interrupt request occurs, if the
system is interruptable, i.e. the interrupt mask is cleared, the
handler function $schedule()$ will be invoked; otherwise $schedule()$
will be pending until the interrupt mask becomes cleared.  The
implementation is shown as $schedule()$ in Algorithm~\ref{a:schedule},
where $taskList$ is the list of periodic tasks to be scheduled. We
assume that the list is ordered descendingly by priorities, and both
variables $taskList$ and $timer$ are global. In this implementation,
there is only one kind of interrupt, the period $T_i$ of each task is
a multiple of $T$, and the tasks are independent, meeting
assumption~(A2).

@@

In Algorithm~\ref{a:schedule}, the handler function $scheduler()$
first updates status of all tasks in $taskList$ via function
$updateStatus()$. Then it traverses the list to execute the ready
tasks one by one, or to do a return when encountering an
interrupted\footnote{Note that the status \textit{INTERRUPT} indicates
  the task is interrupted for the moment, or was interrupted before
  but its execution is not complete yet.} task. The function
$updateStatus()$ updates each task in two steps: firstly, if the task
is running, it becomes interrupted; secondly for the task at its
initiation time, if its previous job is complete, it would be set to
be ready, otherwise it misses its deadline, producing an error. Notice
that $schedule()$ is invoked only when the periodic interrupt is
handled. Its execution cannot be interrupted when it is updating
status of tasks or finding the next task to execute, however, it can
be interrupted while executing some task (Line~\ref{l:function}). This
allows the execution of $schedule()$ to be nested.

In the rest of this paper, we use \emph{scheduling} to refer to the
stage, from the moment when a pending interrupt request is detected,
to the moment when the first should-be-run periodic task starts
executing, i.e. Line~\ref{l:function} in
Algorithm~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when an interrupt request
is handled, (ii) the time spent by $schedule()$ finding and setting
the \emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}), and (iii) the time for switching context
from $schedule()$ to that first should-be-run task. \emph{Switching}
refers to the stage, from the moment when a periodic task completes
its execution, to the moment when the next should-be-run periodic task
starts executing. \emph{Switching time} thus also consists of three
parts: (i) the time for switching context from the finished task back
to $schedule()$, (ii) the time spent by $schedule()$ finding and
setting the \emph{next} should-be-run periodic task, and (iii) the
time for switching context from $schedule()$ to that next
should-be-run task.

@@

These assumptions make our model different from the standard one. For
instance, if an interrupt request occurs while switching is performed,
jobs corresponding to task $\tau_i$ may not be initiated at exactly
$kT_i$. They should wait until switching finishes and the interrupt
mask is cleared. Note that (A2) says that jobs are ready to run at
their initiation times, however, no one can start running at exactly
its initiation time, because scheduling takes time.

\jx{TODO. A figure is needed here to demonstrate the scheduling under
  the three assumptions above.}

In this section, we introduce first how we model the states--the
static aspect--of the system using algebraic data types, then how we
specify the essential behaviors--the dynamic aspect--using rewrite
rules. In particular, modeling of instaneous behaviors would be
explained in Sections~\ref{ss:ir}, \ref{ss:init}
and~\ref{ss:inthandling}, followed by timed behaviors in
Section~\ref{ss:timedbehavior}.

\subsection{Basic Data Types}
In our model, tasks are identified by their indexes of sort \verb|Nat|
in the $taskList$. We define a sort \verb|MaybeNat| wrapping
\verb|Nat|s to refer to some task, with \emph{constructor} \verb|some|
followed by a \verb|Nat| $n$ indicating the task indexed $n$, and
\verb|none| for no task:
\begin{verbatim}
  op none : -> MaybeNat [ctor] .
  op some_ : Nat -> MaybeNat [ctor] .
\end{verbatim}

@@

We also need a sort \verb|Interval|
to denote computation requirements of tasks, 
and a sort \jx{} \verb|Counter| to
represent the time some task has been running and the corresponding
possible finishing interval.
\begin{verbatim}
  op [_,_] : Time Time -> Interval [ctor] .
  op [_/_] : Time Interval -> Counter [ctor] .
\end{verbatim}

@@

\subsection{Interrupt Requests}
\label{ss:ir}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instaneous action, and thus is modeled by the
following instaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the operator \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
flag indicating there is an interrupt request to be handled:
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq = [ (REGS).setIr ; S ] .
\end{verbatim}
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Algorithm~\ref{a:schedule}, which is modeled by
the recursive operation as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and \verb|update_with_| updating the status of individual task
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Algorithm~\ref{a:schedule})
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, status : ST > with TIMER
        = if ST == DORMANT then < O : PTask | status : READY >
          else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > with TIMER
       = if ST == RUNNING then < O : PTask | status : INTERRUPT >
         else < O : PTask |> fi [owise] .
\end{verbatim}
In the case where \verb|TIMER|($timer$) can be divided by the task's
period, i.e. the task should be initiated, if the status is
\verb|DORMANT|, it should be ready to run; otherwise it means the
previous job of the task is not finished, thus it misses its deadline,
producing an \verb|error|. In the other case where the tasks should
not be initiated, the status of the task changes only if it is
\verb|RUNNING|. We can see that \verb|updateStatus_with_| behaves the
same as $updateStatus()$.

@@

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
\emph{Timed behaviors} of the system consist of two parts: the
execution of tasks and the execution of the interrupt source. Both are
modeled into the following one ``standard'' tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instaneous
transition \emph{must} be performed. Notice that the variable $R$ is
\emph{continuous} with respect to the specific time
domain\footnote{Real-Time Maude contains built-in modules to define
  the time domain to be natural numbers and rational numbers,
  specifying \emph{discrete} time domains and \emph{dense} time
  domains respectively.}  that we choose to instantiate the model on,
which is different from timed automata that discretize dense time by
defining ``clock region''.
